Arrows for Parallel Computations

### 2017-11-28 paper/content/motivation.tex
minor
reworking contributions

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common \ptcomment{quantify, e.g. less than 10 \%} \mbcomment{we do this in the contributions section of the Introduction, enough?}, very low-penalty programming interface that allows to switch
the parallel backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

\paragraph{Contributions}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
We propose a new Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]| in Section \ref{sec:parallel-arrows}, which converts a list of Arrows into a new parallel Arrow. Because of this, we do not lose any benefits of using Arrows as the parallelism is encapsulated as another combinator instead of a different type. The resulting Arrow can be used in the same way as a potential serial version.

This is a big advantage as we do not introduce any new types as Monad solutions like the |Par| Monad do. We can just \enquote{plug} in parallel parts into sequential Arrow-based programs and libraries without having to change integral parts of the code. \mbcomment{mention Functions here? We have everything Eden has, but more.}

@@

We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic. This way, we can have multiple backends -- currently a GpH, a |Par| Monad and an Eden version are supported. The backends can easily be swapped, so we are not bound to any specific one. So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.

We introduce a |Future| type class in Section \ref{sec:futures} to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|). Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node. As the |Par| Monad and GpH backends (in their current form) only support shared memory execution and do not have to fix any communication problems, we only use a wrapping dummy |BasicFuture|.

We show that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}) and finally demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced in Section \ref{sec:benchmarks}, i.e. less than $1.7\%$ in cases outside of the error bar of the measurement. \mbcomment{fix this after benchmarks are finally done}.

>>>>

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common \ptcomment{quantify, e.g. less than 10 \%} \mbcomment{we do this in the contributions section of the Introduction, enough?}, very low-penalty programming interface that allows to switch
the parallel backends at will. The induced penalty was max. 8.3 \% in our measurements. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

\paragraph{Contributions}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
We propose a new Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]| in Section \ref{sec:parallel-arrows}, which converts a list of Arrows into a new parallel Arrow. A parallel Arrow is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version.

We evaluate the expressive power of Arrow formalism in the context of parallel programming.

\begin{itemize}
\item We introduce a parallel evaluation formalism using Arrows. One big advantage is that we do not introduce any new types (Sec.~\ref{sec:parallel-arrows}) in contrast Monad solutions such as |Par| Monad. This behaviour encourages better composability.
\item We utilitze multiple backends -- currently a GpH, a |Par| Monad, and Eden. We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic. The backends can easily be swapped, so we are not bound to any specific one.

So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant. (Sec.~\ref{sec:parallel-arrows})
\item We extend our PArrows formalism with |Future|s. Our goal here is to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|). Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node. (Sec.~\ref{sec:futures})
\item It is possible to define algorithmic skeletons with PArrows (Sec.~\ref{sec:skeletons})
\item Finally, we demonstrate that Arrow parallelism is a viable alternative to existing approaches. It introduces only low overhead (Sec.~\ref{sec:benchmarks}).
\end{itemize}

@@
